\section{Introduction}
\label{sec:intro}

\smyrna\ is an application for viewing graphs. It allows the user
to open a window on a graph; navigate around the graph using pan
and zoom; and select nodes and edges with the mouse. The GUI provides various simple
mechanisms for changing the view of the graph and querying information about nodes and edges. 
In addition, \smyrna\
allows the user to perform sophisticated querying and
manipulation of graphs.

Although it works with graphs of any size, \smyrna\ was 
designed to handle large graphs, on the order of 100,000 nodes
and edges. It uses the OpenGL library which allows it to take advantage 
of modern video cards' graphics rendering features. 

\subsection{File Format} 
\label{sec:files}
Currently, \smyrna\ supports only the Graphviz \DOT\ language.
Please refer to 
\begin{center}
\url{http://www.graphviz.org/Documentation.php} 
\end{center}
to get more information about the \DOT\ language of Graphviz and
the related attributes.

\smyrna\ assumes the input graph has position information
for all nodes. This is supplied by the {\tt pos} attribute, whose
format is two or three numbers separated by comma. Thus, for a
node {\tt start}, one might have
\begin{center}
{\tt start [pos="23.5,288"]}
\end{center}
If a graph is laid out using one of the Graphviz graph drawing programs,
the position information is attached in this manner. For example, running the
command
\begin{center}
{\tt sfdp -o outgraph.gv ingraph.gv}
\end{center}
will lay out the graph described in {\tt ingraph.gv} and create the file
{\tt outgraph.gv}, which can be used as input to \smyrna.

All \smyrna\ settings are stored as graph attributes.
For example, you can change the background color by setting the
{\tt bgcolor} attribute of the graph. Or the size of a 
node can be altered by changing its {\tt size} attribute. 
These attributes can be modified through the \smyrna\ GUI, 
using the built-in scripting (cf.~\ref{sec:gvpr}),
or by simply changing the graph file itself. 

Normally in \smyrna, nodes are represented by filled disks, which
can vary in size and color. For smaller graphs, \smyrna\ 
can render the more complex shapes and styles available in Graphviz.
This requires the nodes and edges in the graph to have the necessary
{\it xdot} attributes to describe the rendering. This can be achieved
by running the Graphviz layout with the {\tt -Txdot} flag, e.g.
\begin{center}
{\tt sfdp -Txdot -o outgraph.gv ingraph.gv}
\end{center}
For more information about {\it xdot} attributes and their formats,
see
\begin{center}
\url{http://www.graphviz.org/doc/info/output.html#d:xdot }
\end{center}

\subsection{Visualization Modes} 
\label{sec:vizmodes}
{\it Smyrna} has three display modes:
2D, 3D and topological fisheye. 
In 2D mode, the graph is displayed in the plane. Navigation is
limited to panning and zooming. Any third coordinate in a
node's {\tt pos} attribute is ignored.
In 3D mode, the graph is displayed using 3D graphics, allowing the
user to rotate about the graph in addition to panning and zooming.
Any node whose {\tt pos} attribute does not have a third coordinate
will be placed in the xy plane.

The topological fisheye mode~\cite{topfish} presents a 
distorted, fisheye view of the
graph, in which the areas of the graph near to one or more focus nodes
are shown accurately, while the areas of the graph away from the foci
become more ``compressed'' the further they are from the foci.
For large graphs, this allows the user to see areas of interest clearly
and in detail, while eliding detail away from the foci.
Unlike ordinary fisheye techniques, in which
the distortion is done geometrically, with remote nodes being
drawn on top of each other, 
topological fisheye is a specialized type of semantic zoom in which
remote nodes are combined into meta-nodes with induced edges, thereby
maintaining the graph's topology in a coarsened form.

\subsection{Querying, Filtering and Modifying Graphs}   
Beyond the simple viewing and navigating around a graph, \smyrna\
also provides machinery for querying properties of the graph, modifying
the graph and creating new graphs. The main engine for this is \gvpr,
a scripting language specifically designed to edit graphs. By using 
\gvpr, you can do almost anything related to graphs, from counting 
the number of nodes with an attribute matching a given regular expression to
fairly complex graph algorithms.

One can use \gvpr\ to read and write files. In particular, \gvpr\ scripts
can write to {\tt stdout}. This is implemented in \smyrna\ but a console
window, which is part of the \gvpr\ control tab (Section~\ref{sec:gvprtab}).

When \gvpr\ changes the structure or attributes of the active graph,
these changes are reflected in \smyrna. Conversely, nodes and edges
that are selected have an attribute {\tt selected=1}, which \gvpr\
can use to provide actions on selected nodes and edges.

The use of \gvpr\ is discussed in more detail in Section~\ref{sec:gvpr}.

